Type Evolution in Interactive Programming Systems

### 2024-05-31 paper.tex
WIP tweaking statemachines

\usepackage{csquotes}
\usepackage{changepage}
\usepackage{multicol}
\usepackage{ccicons}
\usepackage[many]{tcolorbox}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{framed}

@@

@@

\begin{comment}
\DeclareRobustCommand{\frameworkbox}[2][gray!15]{
\begin{tcolorbox}[breakable,left=3pt,right=3pt,top=3pt,bottom=3pt,colback=#1,colframe=#1,parbox=false,
  width=\dimexpr\textwidth\relax,enlarge left by=0mm,boxsep=5pt,arc=0pt,enlarge top by=0.5em,%enlarge bottom by=0.0em,
  outer arc=0pt]\setlength{\parskip}{0.5em}#2
\end{tcolorbox}}
\newcommand{\frameworkboxtitle}[1]{\setlength{\parindent}{0em}{\firamedium #1.}\quad}
\end{comment}

@@

\section{Divergence Control in Spreadsheets and Databases}
Type evolution becomes more challenging when data, code, and types cannot all be updated
together in an atomic way. In these situations, different versions must coexist and interoperate
with one another. For example, edits done to data of a new type must be applied to
data of an older type, or code written against older type must run against data of a newer type.

@@

%\newpage

\section{Live Modeling Languages Require Run-Time State Migration}
\label{livemodeling}
%Related work
%``Slogan'': editing a program is diverging from run-time schema.

\subsection{Context}
In stateful live programming, a change in the program can make the run-time state of the execution out of date, hence we want to migrate the run-time state to keep on running.

One can see a program itself as an instance of a static schema (its AST type), that in turn determines (defines/implies/induces) a run-time schema: the run-time structures of code (e.g., classes, inheritance links, declarations, method definitions etc.), as well as the structure of the run-time state. Whereas the run-time \textit{code} structures do not typically change while using an application, the run-time \textit{state} constantly changes.
Live programming, however, requires reconciling changes to the program with the run-time structures of both the code and the state. This typically means that at a certain point (a quiescent point) during execution, the code structures need to be updated (hot swapped), and the state needs to be \textit{migrated}.

@@

\begin{figure}[t]
\centering
\begin{minipage}[t]{0.4\textwidth}
\begin{lstlisting}[language=java,morekeywords={machine,on,state,var}]
machine Doors

var isClosed: bool = true

state closed
  isClosed := true
  on open => opened

state opened
  isClosed := false
  on close => closed
\end{lstlisting}
\end{minipage}
\hspace*{5pt}\vline
\begin{minipage}[t]{0.4\textwidth}
\begin{lstlisting}[language=java,morekeywords={on}]
  class Machine
    name: string
    states: State*
    vars: Decl*

  class State
    name: string
    onEntry: Stmt*
    transitions: Trans*

  class Trans
    event: string
    target: State

  class Decl
    name: string
    type: Type
  \end{lstlisting}
\end{minipage}
\caption{A statemachine and its abstract syntax schema (omitting \lstinline{Stmt} and \lstinline{Type})}
\label{LST:statemachines}
\end{figure}

@@

\paragraph{Live programming.}
Live programming~\cite{tanimoto90, rein2018exploratory} seeks to speed up feedback in (primarily solo) programming.
% \cite{mcdirmid13, Rauch_2019, beckmann2021shortening}
\emph{Hot reloading/swapping}~\cite{barenz2020essence, hicks2005dynamic} installs code changes into a running system preserving some runtime state but not attempting to migrate that state through type changes.
\citet{SemanticDeltas} proposes a research agenda for thoroughly live DSL programming focusing on \emph{Semantic Deltas} that unify edit-time and run-time change.
\citet{RuntimeConstraint} study the live modeling problem in Section \ref{livemodeling}. They use constraints supplied by the developer to solve for the best new runtime state following a DSL language evolution. \citet{vanRozen19} take an alternative approach to the same problem using \emph{origin tracking} in textual editing.
%\citet{vanRozen23} subsequently proposed a metaprogramming language for live DSL programming.

>>>>

\usepackage{csquotes}
\usepackage{changepage}
\usepackage{multicol}
\usepackage{xcolor}
\usepackage{ccicons}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{framed}

@@

% \definecolor{progblue}{RGB}{64,104,224}
% \renewenvironment{leftbar}[1][\hsize]
% {%
%     \def\FrameCommand
%     {%
%         {\color{progblue}\vrule width 3pt}%
%         \hspace{0pt}%must no space.
%         \fboxsep=\FrameSep\colorbox{white}%
%     }%
%     \MakeFramed{\hsize#1\advance\hsize-\width\FrameRestore}%
% }
% {\endMakeFramed}

@@

@@

\section{Divergence Control in Spreadsheets}
Type evolution becomes more challenging when data, code, and types cannot all be updated
together in an atomic way. In these situations, different versions must coexist and interoperate
with one another. For example, edits done to data of a new type must be applied to
data of an older type, or code written against older type must run against data of a newer type.

@@

% ==================================================================================================
% MODELING
% ==================================================================================================

\begin{figure}[t]
\begin{adjustwidth}{-1.5em}{-1.5em}
\begin{minipage}[t]{0.26\textwidth}
\begin{lstlisting}[language=java,morekeywords={on},mathescape,numbers=none]
class Machine
  name: string
  states: State*
  vars: Decl*
\end{lstlisting}
\end{minipage}
\begin{minipage}[t]{0.28\textwidth}
\begin{lstlisting}[language=java,morekeywords={on},mathescape,numbers=none]
class State
  name: string
  onEntry: Stmt*
  transitions: Trans*
\end{lstlisting}
\end{minipage}
\begin{minipage}[t]{0.24\textwidth}
\begin{lstlisting}[language=java,morekeywords={on},mathescape,numbers=none]
class Trans
  event: string
  target: State
$~$
\end{lstlisting}
\end{minipage}
\begin{minipage}[t]{0.24\textwidth}
\begin{lstlisting}[language=java,morekeywords={on},mathescape,numbers=none]
class Decl
  name: string
  type: Type
$~$
\end{lstlisting}
\end{minipage}
\end{adjustwidth}
\vspace{-1em}
\caption{A statemachine and its abstract syntax schema (omitting \lstinline{Stmt} and \lstinline{Type})}
\label{lst:machinedef}
\vspace{0.5em}
\end{figure}

\section{Live Modeling Languages}
\label{sec:livemodeling}
%Related work
%``Slogan'': editing a program is diverging from run-time schema.

We considered live programming in the context of the Elm architecture in Section~\ref{sec:elm},
where the challenge was to migrate run-time state (data) to a new type to keep the program running.
Work on language modeling offers a new perspective on the problem. In systems such as Rascal
\cite{klint2011easy}, one can see a program in a domain-specific language not just as \emph{types}
and \emph{code}, but also as \emph{data}, or an instance of a metamodel (a static type) that
defines the AST of the domain-specific language. The program in a domain-specific language
that in turn determines (defines/implies/induces) a run-time schema: the run-time structures
of code (e.g., classes, inheritance links, declarations, method definitions etc.), as well as
the structure of the run-time state.

In the absence of live programming, the run-time \emph{code} structures do not typically
change while using an application and the run-time \emph{state} changes constantly.

Live programming, however, requires reconciling changes to the program with the run-time structures
of both the code and the state. This typically means that at a certain point (a quiescent point)
during execution, the code structures need to be updated (hot swapped), and the state needs to be \textit{migrated}.

@@

\noindent
\begin{adjustwidth}{-0.5em}{0em}
\begin{minipage}[t]{0.32\textwidth}
\begin{lstlisting}[language=java,morekeywords={machine,on,state,var},numbers=none]
machine Doors

var isClosed: bool = true
\end{lstlisting}
\end{minipage}
\begin{minipage}[t]{0.32\textwidth}
\begin{lstlisting}[language=java,morekeywords={machine,on,state,var},numbers=none]
state closed
  isClosed := true
  on open => opened
\end{lstlisting}
\end{minipage}
\begin{minipage}[t]{0.32\textwidth}
\begin{lstlisting}[language=java,morekeywords={machine,on,state,var},numbers=none]
state opened
  isClosed := false
  on close => closed
\end{lstlisting}
\end{minipage}
\end{adjustwidth}

%\begin{figure}[t]
%\centering
%\caption{A statemachine and its abstract syntax schema (omitting \lstinline{Stmt} and \lstinline{Type})}
%\label{LST:statemachines}
%\end{figure}

@@

\paragraph{Live programming.}
Live programming~\cite{tanimoto90, rein2018exploratory} seeks to speed up feedback in (primarily solo) programming.
% \cite{mcdirmid13, Rauch_2019, beckmann2021shortening}
\emph{Hot reloading/swapping}~\cite{barenz2020essence, hicks2005dynamic} installs code changes into a running system preserving some runtime state but not attempting to migrate that state through type changes.
\citet{SemanticDeltas} proposes a research agenda for thoroughly live DSL programming focusing on \emph{Semantic Deltas} that unify edit-time and run-time change.
\citet{RuntimeConstraint} study the live modeling problem in Section \ref{sec:livemodeling}. They use constraints supplied by the developer to solve for the best new runtime state following a DSL language evolution. \citet{vanRozen19} take an alternative approach to the same problem using \emph{origin tracking} in textual editing.
%\citet{vanRozen23} subsequently proposed a metaprogramming language for live DSL programming.
